---
title: "2025_reu_hypothesis_2_dbh"
output: 
  html_document:
          toc: true
          toc_depth: 4
date: "2025-07-15"
---


# Hypothesis 2: Higher DBH will be correlated with a higher percentage of canker on the trunk and base
* Variables: DBH→ canker_area_base/trunk
* Mechanism: a large, healthy tree might fight off the canker for longer amount of time while small trees are the opposite

# Data Import
## Importing 2024 & 2025 Data
See the full code for the break down for the importing and cleaning open the documents of the same name. 

### Create the R files if needed. 
First, I need to retrieve only the R code from the cleaning files. These files also include text descriptions of what cleaning is done, which is not needed for importing the code to do the data cleaning here. 
```{r EVAL=FALSE, IGNORE=TRUE}
# purl("G:/My Drive/1 Projects/REU2425 - Morton Arboretum/R/REU2025-data-exploration/week8_cleaning_2024_health_form_data.Rmd")
# 
# purl("G:/My Drive/1 Projects/REU2425 - Morton Arboretum/R/REU2025-data-exploration/week8_cleaning_2025_health_form_data.Rmd")
```


### Run the R files to get cleaned data
Next, we can run those extracted R files to actually import and clean the data. 
```{r message=FALSE, warning=FALSE}
## 2024 data
source("G:/My Drive/1 Projects/REU2425 - Morton Arboretum/R/REU2025-data-exploration/week8_cleaning_2024_health_form_data.R")

## 2025 data
source("G:/My Drive/1 Projects/REU2425 - Morton Arboretum/R/REU2025-data-exploration/week8_cleaning_2025_health_form_data.R")
```


# Setup of Analysis

This chunk defines a re-usable function to run a regression across multiple input datas. 

```{r define_hypoth2_analysis}
do_hypoth2_analysis <- function(data, predictor, response, xlabel, ylabel, color = "black") {
  # Define linear model
  model <- lm(reformulate(predictor, response), data = data)
  coefs <- coef(model)
  r_squared <- summary(model)$r.squared
  
  # Create equation and R^2 text
  equation <- paste0("y = ", round(coefs[2], 2), "x + ", round(coefs[1], 2))
  r2_text <- paste0("italic(R)^2 == ", round(r_squared, 4))
  
  # Get plotting ranges
  max_x <- max(data[[predictor]], na.rm = TRUE)
  
  max_y <- max(data[[response]], na.rm = TRUE)
  min_y <- min(data[[response]], na.rm = TRUE)
  range_y <- max_y - min_y
  
  # Annotate sample size near the bottom-right
  sample_size_y_position <- min_y + 0.08 * range_y
  
  # Annotate equation and r2 near the top-right
  eq_y_position <- max_y - 0.05 * range_y
  r2_y_position <- max_y - 0.12 * range_y
  
  # PLOT ------------------------------------------------------
  ggplot(data, aes_string(x = predictor, y = response)) +
    # Points
    geom_point(color = color) +
    
    # Line of best fit
    geom_smooth(method = "lm", color = color) +
    
    # Axes Label & Theme
    labs(x = xlabel, y = ylabel) +  # Set the y-axis label
    theme(
      axis.title.y = element_text(color = color) # Change y-axis label color
    ) + 
    
    # Samples size text
    annotate(
      "text", 
      x = max_x,
      y = sample_size_y_position,
      label = paste("n = ", count(data)),
      hjust = 1,
      size = 6,
      color = "black"
    ) +
    
    # Equation and R^2 text
    annotate(
      "text",
      x = max_x,
      y = eq_y_position,
      label = equation,
      hjust = 1,
      size = 4,
      color = color
    ) +
    annotate(
      "text",
      x = max_x,
      y = r2_y_position,
      label = r2_text,
      hjust = 1,
      size = 4,
      color = color,
      parse = TRUE # Ensures that italics styling works
    )
}
```


# Data Analysis
Here are the major delinations of questions across this hypotheses:

* What patterns of DBH and canker areas do we see overall-- across all sites, years and age classes?
* What patterns of DBH and canker areas do we see across years?
* What patterns of DBH and canker areas do we see across seedlings?
* What patterns of DBH and canker areas do we see across adults?
* What patterns of DBH and canker areas do we see across sites?

Notably: 
* I'd like to include both base and trunk canker area on the same graph; I think that could be very nice

## Patterns overall-- across all individuals (all sites, years, and age classes)
```{r}

# Combining 2024 and 2025 data along canker area & DBH
#combined_2024_2025 <- 
```



## Patterns across years
### Patterns across 2025
```{r run_2025_analysis, message=FALSE, warning=FALSE}
library(scales)

hex <- hue_pal()(3)

patterns_across_2025_base <- do_hypoth2_analysis(
  data = health_assess_2025, 
  predictor = "dbh_cm", 
  response = "base_canker_area",
  xlabel = "All 2025 Individuals DBH (cm)",
  ylabel = "% of Base Canker Area",
  color = hex[1]
)
patterns_across_2025_base

patterns_across_2025_trunk <- do_hypoth2_analysis(
  data = health_assess_2025, 
  predictor = "dbh_cm", 
  response = "trunk_canker_area",
  xlabel = "All 2025 Individuals DBH (cm)",
  ylabel = "% of Trunk Canker Area",
  color = hex[2]
)
patterns_across_2025_trunk

patterns_across_2025_girdled <- do_hypoth2_analysis(
  data = health_assess_2025, 
  predictor = "dbh_cm", 
  response = "girdled_canker_circum_2025",
  xlabel = "All 2025 Individuals DBH (cm)",
  ylabel = "% of Circumference Girdled",
  color = hex[3]
)
patterns_across_2025_girdled
```

### Patterns across 2024
```{r message=FALSE, warning=FALSE}
library(scales)

hex <- hue_pal()(3)

patterns_across_2024_base <- do_hypoth2_analysis(
  data = health_assess_2024, 
  predictor = "dbh_cm", 
  response = "base_canker_area",
  xlabel = "All 2024 Individuals DBH (cm)",
  ylabel = "% of Base Canker Area",
  color = hex[1]
)
patterns_across_2024_base

patterns_across_2024_trunk <- do_hypoth2_analysis(
  data = health_assess_2024, 
  predictor = "dbh_cm", 
  response = "trunk_canker_area",
  xlabel = "All 2024 Individuals DBH (cm)",
  ylabel = "% of Trunk Canker Area",
  color = hex[2]
)
patterns_across_2024_trunk
```


## Patterns across sites
**************

## Results
### Girdled circumference: Negative correlation between DBH and % circumference girdled
This result would indicate that younger trees have slightly higher girdling. 

We may be seeing this trend due to an inaccuracy with attempting to estimate circumference on smaller trees.
```{r message=FALSE}
patterns_across_2025_girdled
```

### 2024 versus 2025: Appears to be an increase in DBH ~ Canker Area over 2024 and 2025.
```{r message=FALSE}
library(patchwork)

(patterns_across_2024_base + patterns_across_2025_base) + plot_annotation("% Base Canker Area 2024 versus 2025")

(patterns_across_2024_trunk + patterns_across_2025_trunk) + plot_annotation("% Trunk Canker Area 2024 versus 2025")
```
  
  
# Future work

## Comparing base and trunk areas on same graph
## Two response variables (base/trunk compared together)
```{r eval=FALSE}
# 
# # This ensures adds a DBH measure to each of the canker entries, such that they can be plotted against each other. 
# data <- health_assess_2025
# predictor <- "dbh_cm"
# response1 <- "base_canker_area"
# response2 <- "trunk_canker_area"
# 
# long_data <- data %>%
#   select(all_of(c(predictor, response1, response2))) %>%
#   pivot_longer(
#     cols = c(all_of(response1), all_of(response2)),
#     names_to = "response_type",
#     values_to = "response_value"
#   )
# 
# # Fit models and extract stats for annotation
# annotations <- long_data %>%
#   group_by(response_type) %>% # Collects all of the base/trunk canker area readings together
#   group_modify(~ {
#     model <- lm(response_value ~ get(predictor), data = long_health_data)
#     coefs <- coef(model)
#     r_squared <- summary(model)$r.squared
#     eq <- paste0("y == ", round(coefs[2], 2), " * x + ", round(coefs[1], 2))
#     r2 <- paste0("italic(R)^2 == ", round(r_squared, 4))
#     data.frame(equation = eq, r2 = r2)
#   })
# 
# # Determine top-right positions for annotation
#   max_x <- max(long_data[[predictor]], na.rm = TRUE)
#   max_y <- max(long_data$response_value, na.rm = TRUE)
#   min_y <- min(long_data$response_value, na.rm = TRUE)
#   range_y <- max_y - min(long_data$response_value, na.rm = TRUE)
# 
#   # Add y positions for annotations
#   # annotations["base_canker_area"]$eq_y <- 
#   annotations$eq_y <- min_y + 0.05 * range_y
#   annotations$r2_y <- min_y + 0.12 * range_y
#   annotations$x <- max_x
# 
#   # Plot
#   ggplot(long_data, aes_string(x = predictor, y = "response_value", color = "response_type")) +
#     geom_point() +
#     geom_smooth(method = "lm", se = FALSE) +
#     # xlab(xlabel) +
#     # ylab(ylabel) +
#     # Add annotations for each response type
#     geom_text(data = annotations,
#               aes(x = x, y = eq_y, label = equation, color = response_type),
#               hjust = 1, size = 4, parse = TRUE, inherit.aes = FALSE) +
#     geom_text(data = annotations,
#               aes(x = x, y = r2_y, label = r2, color = response_type),
#               hjust = 1, size = 4, parse = TRUE, inherit.aes = FALSE) +
#     theme_minimal()
# 
# view(annotations)
```
