---
title: "2025_WCP_site_data_exploration"
output: 
  html_document:
          toc: true
          toc_depth: 4
date: "2025-07-17"
---

EXPLORING WCP DATA ONLY

# Importing 2025 Data
See the full code for the break down for the importing and cleaning open the documents of the same name. 

## Create the R files if needed. 
First, I need to retrieve only the R code from the cleaning files. These files also include text descriptions of what cleaning is done, which is not needed for importing the code to do the data cleaning here. 
```{r IGNORE=TRUE}

# purl("cleaning_2025_health_form_data.Rmd",
#      output =
#        paste("purl/cleaning_2025_health_form_data",
#        Sys.Date(),".R"))
```


## Run the R files to get cleaned data
Next, we can run those extracted R files to actually import and clean the data. 
```{r message=FALSE, warning=FALSE}
## 2025 data
source(paste("purl/cleaning_2025_health_form_data", Sys.Date(), ".R"))
```


# Select only the WCP data
```{r}
health_assess_2025 <- health_assess_2025 %>% slice(1:66)
```

# Data exploration
* Need to include girdled circum

## Proportion of seedlings
* NAs are from the first few days where we didn't include the seedling question yet; we could adjust these
```{r}
bar_seedlings <- health_assess_2025 %>% ggplot(aes(x = seedling_y_n)) +
  geom_bar(aes(fill = seedling_y_n)) + 
  xlab("2025 WCP Seedlings Y/N")

bar_seedlings
```

## Plant Height
** Would love to fix the seedling/adult coloring

### All individual distribution
```{r message=TRUE, warning=TRUE, paged.print=TRUE}
library(patchwork)

# Plant Height (ft)
mean_plant_height <- mean(health_assess_2025$plant_height_ft, na.rm = TRUE)
median_plant_height <- median(health_assess_2025$plant_height_ft, na.rm = TRUE)
print(mean_plant_height)

hist_plant_height <- health_assess_2025 %>% ggplot(aes(x = health_assess_2025$plant_height_ft)) +
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) +
  xlab("2025 WCP All Individuals")
  ylab("Number of Individuals") +
  # Mean
  geom_vline(
    aes(xintercept = mean_plant_height),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 17,
           y = 15,
           label = paste("Mean =", round(mean_plant_height, 2), "ft")) +

  # Median
  geom_vline(
    aes(xintercept = median_plant_height),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 11,
           y = 20,
           label = paste("Median =", round(median_plant_height, 2), "ft")) 

  hist_plant_height
```

### Seedling subset
```{r message=TRUE, warning=TRUE, paged.print=TRUE}
seedlings_2025 <- health_assess_2025 %>% filter(is.na(dbh_cm))


# Plant Height (ft)
mean_plant_height <- mean(seedlings_2025$plant_height_ft, na.rm = TRUE)
median_plant_height <- median(seedlings_2025$plant_height_ft, na.rm = TRUE)
print(mean_plant_height)
hist_plant_height_seedling <- seedlings_2025 %>% ggplot(aes(x = plant_height_ft)) +
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) +
  xlab("2025 WCP Seedling Height (ft)")
  ylab("Number of Individuals") +
  # Mean
  geom_vline(
    aes(xintercept = mean_plant_height),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 3.5,
           y = 5,
           label = paste("Mean =", round(mean_plant_height, 2), "ft")) +

  # Median
  geom_vline(
    aes(xintercept = median_plant_height),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 3,
           y = 5,
           label = paste("Median =", round(median_plant_height, 2), "ft"))
hist_plant_height_seedling
```


### Adult subset
```{r message=TRUE, warning=TRUE, paged.print=TRUE}
adult_2025 <- health_assess_2025 %>% filter(is.na(dbh_cm))


# Plant Height (ft)
mean_plant_height <- mean(adult_2025$plant_height_ft, na.rm = TRUE)
median_plant_height <- median(adult_2025$plant_height_ft, na.rm = TRUE)
print(mean_plant_height)


hist_plant_height_adult <- adult_2025 %>% ggplot(aes(x = plant_height_ft)) +
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) +
  xlab("2025 WCP Adult Height (ft)")
  ylab("Number of Individuals") +
  # Mean
  geom_vline(
    aes(xintercept = mean_plant_height),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 3.5,
           y = 5,
           label = paste("Mean =", round(mean_plant_height, 2), "ft")) +

  # Median
  geom_vline(
    aes(xintercept = median_plant_height),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 3,
           y = 5,
           label = paste("Median =", round(median_plant_height, 2), "ft"))
hist_plant_height_adult
```




## DBH

### All 2025 Individuals:
```{r}
# DBH (cm)
mean_dbh <- mean(health_assess_2025$dbh_cm, na.rm = TRUE)
median_dbh <- median(health_assess_2025$dbh_cm, na.rm = TRUE)

# violin_dbh <- health_assess_2025 %>% ggplot(aes(x = dbh_cm)) + 
#   geom_violin()
# 
# violin_dbh

hist_dbh <- health_assess_2025 %>% ggplot(aes(x = dbh_cm)) +
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) + 
  xlab("2025 All *WCP* Individuals DBH (cm)")
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_dbh),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 30,
           y = 5,
           label = paste("Mean =", round(mean_dbh, 2), "cm")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_dbh),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 20,
           y = 8,
           label = paste("Median =", round(median_dbh, 2), "cm")) 
hist_dbh
```



### Adults: Most adults in WCP were saplings. 
```{r}
adults_2025 <- health_assess_2025 %>% filter(!is.na(dbh_cm))

# DBH (cm)
mean_dbh <- mean(adults_2025$dbh_cm, na.rm = TRUE)
median_dbh <- median(adults_2025$dbh_cm, na.rm = TRUE)
hist_dbh <- adults_2025 %>% ggplot(aes(x = dbh_cm)) +
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) + 
  xlab("2025 WCP Adult DBH (cm) [Should only be adults]")
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_dbh),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 30,
           y = 5,
           label = paste("Mean =", round(mean_dbh, 2), "cm")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_dbh),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 20,
           y = 8,
           label = paste("Median =", round(median_dbh, 2), "cm")) 
hist_dbh
```

## % Live Canopy

### All individual distribution
```{r}
mean_percent_live_canopy <- mean(health_assess_2025$percent_live_canopy, na.rm = TRUE)
median_percent_live_canopy <- median(health_assess_2025$percent_live_canopy, na.rm = TRUE)

hist_percent_live_canopy_2025 <- health_assess_2025 %>% 
  ggplot(aes(x = percent_live_canopy)) +
  xlab("% Live Canopy in 2025") +
  ylab("Number of Individuals") + 
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) + 
  # Mean
  geom_vline(
    aes(xintercept = mean_percent_live_canopy),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 70,
           y = 20, 
           label = paste("Mean =", round(mean_percent_live_canopy, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_percent_live_canopy),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 85,
           y = 30,
           label = paste("Median =", round(median_percent_live_canopy, 2), "%")) 

hist_percent_live_canopy_2025
```

### Seedling subset
```{r}
seedlings_2025 <- health_assess_2025 %>% filter(is.na(dbh_cm))

mean_percent_live_canopy <- mean(seedlings_2025$percent_live_canopy, na.rm = TRUE)
median_percent_live_canopy <- median(seedlings_2025$percent_live_canopy, na.rm = TRUE)

hist_percent_live_canopy_2025_seedling <- seedlings_2025 %>% 
  ggplot(aes(x = percent_live_canopy)) +
  xlab("% Live Canopy in 2025 FOR SEEDLINGS") +
  ylab("Number of Individuals") + 
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) + 
  # Mean
  geom_vline(
    aes(xintercept = mean_percent_live_canopy),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 70,
           y = 20, 
           label = paste("Mean =", round(mean_percent_live_canopy, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_percent_live_canopy),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 85,
           y = 30,
           label = paste("Median =", round(median_percent_live_canopy, 2), "%")) 

hist_percent_live_canopy_2025_seedling
```


### Adult subset
```{r}
adults_2025 <- health_assess_2025 %>% filter(!is.na(dbh_cm))

mean_percent_live_canopy <- mean(adults_2025$percent_live_canopy, na.rm = TRUE)
median_percent_live_canopy <- median(adults_2025$percent_live_canopy, na.rm = TRUE)

hist_percent_live_canopy_2025_adult <- adults_2025 %>% 
  ggplot(aes(x = percent_live_canopy)) +
  xlab("% Live Canopy in 2025 FOR ADULTS") +
  ylab("Number of Individuals") + 
  geom_histogram(bins = 30, aes(fill = seedling_y_n)) + 
  # Mean
  geom_vline(
    aes(xintercept = mean_percent_live_canopy),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 70,
           y = 20, 
           label = paste("Mean =", round(mean_percent_live_canopy, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_percent_live_canopy),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 85,
           y = 30,
           label = paste("Median =", round(median_percent_live_canopy, 2), "%")) 

hist_percent_live_canopy_2025_adult

```


<!-- ## Epicormics -->
<!-- ```{r} -->
<!-- # ------------ BASE -->

<!-- mean_hist_epicormic_base <- mean(health_assess_2025$base_epicormics, na.rm = TRUE) -->
<!-- median_hist_epicormic_base <- median(health_assess_2025$base_epicormics, na.rm = TRUE) -->

<!-- hist_epicormic_base <- health_assess_2025 %>% ggplot(aes(x = base_epicormics)) + -->
<!--   geom_histogram(aes(fill = seedling_y_n)) + -->
<!--   ylab("Number of Individuals") +  -->
<!--   geom_histogram(bins = 30, aes(fill = seedling_y_n)) +  -->
<!--   # Mean -->
<!--   geom_vline( -->
<!--     aes(xintercept = mean_hist_epicormic_base), -->
<!--     color = "#bdbdbd", -->
<!--     linetype = "dashed", -->
<!--     size = 1 -->
<!--   ) + -->
<!--   annotate("text", -->
<!--            color = "#bdbdbd", -->
<!--            x = 1.3, -->
<!--            y = 20,  -->
<!--            label = paste("Mean =", round(mean_hist_epicormic_base, 2))) + -->

<!--   # Median -->
<!--   geom_vline( -->
<!--     aes(xintercept = median_hist_epicormic_base), -->
<!--     color = "#636363", -->
<!--     linetype = "solid", -->
<!--     size = 1 -->
<!--   ) + -->
<!--   annotate("text", -->
<!--            color = "#636363", -->
<!--            x = 0.7, -->
<!--            y = 30, -->
<!--            label = paste("Median =", round(median_hist_epicormic_base, 2))) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # ------------ TRUNK -->
<!-- mean <- mean(health_assess_2025$trunk_epicormics, na.rm = TRUE) -->
<!-- median <- median(health_assess_2025$trunk_epicormics, na.rm = TRUE) -->

<!-- hist_epicormic_trunk <- health_assess_2025 %>% ggplot(aes(x = trunk_epicormics)) + -->
<!--   geom_histogram(aes(fill = seedling_y_n)) +  -->
<!--   ylab("Number of Individuals") +  -->
<!--   # Mean -->
<!--   geom_vline( -->
<!--     aes(xintercept = mean), -->
<!--     color = "#bdbdbd", -->
<!--     linetype = "dashed", -->
<!--     size = 1 -->
<!--   ) + -->
<!--   annotate("text", -->
<!--            color = "#bdbdbd", -->
<!--            x = 0.6, -->
<!--            y = 20,  -->
<!--            label = paste("Mean =", round(mean, 2))) + -->

<!--   # Median -->
<!--   geom_vline( -->
<!--     aes(xintercept = median), -->
<!--     color = "#636363", -->
<!--     linetype = "solid", -->
<!--     size = 1 -->
<!--   ) + -->
<!--   annotate("text", -->
<!--            color = "#636363", -->
<!--            x = 0.4, -->
<!--            y = 30, -->
<!--            label = paste("Median =", round(median, 2))) -->

<!-- hist_epicormic_base / hist_epicormic_trunk -->
<!-- ``` -->

## Has Canker & Callous 

```{r}
library(viridis)

health_assess_2025 <- health_assess_2025 %>% mutate(
  purdue_severity_canker = recode(
    health_assess_2025$purdue_severity_canker,
    "1. Fewer than 3 active cankers that are all smaller than 2-3 inches in length or diameter OR fewer than 3 inactive cankers." = "1",
    "2. More than 3 active cankers, OR 2-5 shallow (with no dead tissue) healed over with cracks less than 7 inches long." = "2",
"3. More than 5 active OR inactive cankers cracked through the bark to the tissue below which have healed over, but you still see the level of damage." = "3",
"4. Cankers occur all over the 10-foot area, with deep cracks and both active and inactive cankers." = "4",
"5. Tree almost dead, mostly inactive cankers with deep cracks to dead tissue." = "5"
  )
)

bar_has_canker_2025 <- health_assess_2025 %>% ggplot(aes(x = has_canker)) +
  geom_bar( aes(fill = purdue_severity_canker))

bar_has_callous_2025 <- health_assess_2025 %>% ggplot(aes(x = has_callous)) +
  geom_bar( aes(fill = purdue_severity_canker))

(bar_has_canker_2025 + bar_has_callous_2025) + plot_layout(guides="collect")
```

## Canker Areas


### Box plots
```{r}
health_long <- seedlings_2025 %>%
  mutate(case_id = row_number()) %>%
  pivot_longer(
    cols = ends_with("canker_area"),
    names_to = "canker_type",
    values_to = "canker_percentage"
  ) 

canker_base_box <- health_long %>% ggplot(aes(x = canker_type, y = canker_percentage)) +
  geom_violin(fill = "skyblue", color = "black") +
  geom_point(alpha = 0.2) + 
  geom_jitter(width = 0.1, alpha = 0.5) + 
  theme_minimal() +
  labs(x = "Canker Location", y = "% Covered In Canker")

canker_base_box
```

### Histograms
```{r}
mean_base_area <- mean(health_assess_2025$base_canker_area, na.rm = TRUE)
median_base_area <- median(health_assess_2025$base_canker_area, na.rm = TRUE)

hist_base_canker <- health_assess_2025 %>% ggplot(aes(x = base_canker_area)) +
  geom_histogram( aes(fill = seedling_y_n)) + 
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_base_area),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 26,
           y = 10, 
           label = paste("Mean =", round(mean_base_area, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_base_area),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 10,
           y = 25,
           label = paste("Median =", round(median_base_area, 2), "%"))

```

```{r}
mean_trunk_area <- mean(health_assess_2025$trunk_canker_area, na.rm = TRUE)
median_trunk_area <- median(health_assess_2025$trunk_canker_area, na.rm = TRUE)

hist_trunk_canker <- health_assess_2025 %>% ggplot(aes(x = trunk_canker_area)) +
  geom_histogram(aes(fill = seedling_y_n)) +
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_trunk_area),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 32,
           y = 10, 
           label = paste("Mean =", round(mean_trunk_area, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_trunk_area),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 30,
           y = 13,
           label = paste("Median =", round(median_trunk_area, 2), "%"))

hist_base_canker / hist_trunk_canker
```

```{r}
density_trunk_canker <- health_assess_2025 %>% ggplot(aes(x = trunk_canker_area)) +
  geom_density()
  # + aes(fill = has_callous_2025))
  # # Mean
  # geom_vline(
  #   aes(xintercept = mean_trunk_area),
  #   color = "#bdbdbd",
  #   linetype = "dashed",
  #   size = 1
  # ) +
  # annotate("text",
  #          color = "#bdbdbd",
  #          x = 0.1,
  #          y = 0.1, 
  #          label = paste("Mean =", round(mean_trunk_area, 2), "%")) +
  # 
  # # Median
  # geom_vline(
  #   aes(xintercept = median_trunk_area),
  #   color = "#636363",
  #   linetype = "solid",
  #   size = 1
  # ) +
  # annotate("text",
  #          color = "#636363",
  #          x = 0.5,
  #          y = 0.5,
  #          label = paste("Median =", round(median_trunk_area, 2), "%"))


density_trunk_canker

```

```{r}

# Linear regression model
model <- lm(base_canker_area ~ dbh_cm, data = health_assess_2025)
coefs <- coef(model)
r_squared <- summary(model)$r.squared
equation <- paste("y = ", round(coefs[2], 2), "x", " + ", round(coefs[1], 2),  sep = "")
   
point_base_canker <- health_assess_2025 %>% ggplot(aes(x = dbh_cm, y =
                                                         base_canker_area)) +
  geom_point(aes(color = seedling_y_n)) +
  geom_smooth(method = lm) +
  annotate(
    "text",
    x = 30,
    y = 80,
    label = equation,
    hjust = 0
  ) + 
  annotate(
    "text",
    x = 30,
    y = 65,
    label = paste("R-squared = ", round(r_squared, 4)),
    hjust = 0
  )
```

```{r}
# Linear regression model
model <- lm(trunk_canker_area ~ dbh_cm, data = health_assess_2025)
coefs <- coef(model)
r_squared <- summary(model)$r.squared
equation <- paste("y = ", round(coefs[2], 2), "x", " + ", round(coefs[1], 2),  sep = "")

point_trunk_canker <- health_assess_2025 %>% ggplot(aes(x = dbh_cm, y=trunk_canker_area)) +
  geom_point(aes(color=seedling_y_n)) + 
  geom_smooth(method=lm) + 
    annotate(
    "text",
    x = 30,
    y = 80,
    label = equation,
    hjust = 0
  ) + 
  annotate(
    "text",
    x = 30,
    y = 65,
    label = paste("R-squared = ", round(r_squared, 4)),
    hjust = 0
  )

(point_base_canker / point_trunk_canker) + plot_layout(guides="collect")
```

```{r}
health_assess_2025 %>%
  ggplot() +
  geom_point(aes(x=dbh_cm, y=base_canker_area, colour = purdue_severity_canker)) + 
  theme_classic() + 
  facet_wrap(~purdue_severity_canker)
```


## Canker severity
```{r}
bar_canker_severity <- health_assess_2025 %>% ggplot(aes(x = purdue_severity_canker)) + 
  geom_bar(aes(fill = purdue_severity_canker))

bar_canker_severity
```


## Densiometer

### Box & Violin
```{r}
health_long <- health_assess_2025 %>%
  pivot_longer(cols = starts_with("densio_"),
               names_to = "direction",
               values_to = "percent_cover")


densiometer_box <- health_long %>% ggplot(aes(x = direction, y = percent_cover)) +
  geom_boxplot(fill = "skyblue", color = "black") +
  
  # Conditional fill and color
  geom_boxplot(
    aes(fill = direction == "densio_south", color = direction == "densio_south")
  ) +
  
  geom_jitter(width = 0.1, alpha = 0.5) +

  # Manual scales for TRUE/FALSE
  scale_fill_manual(
    values = c("TRUE" = "tomato", "FALSE" = "skyblue"),
    guide = "none"  # hide fill legend
  ) +
  scale_color_manual(
    values = c("TRUE" = "darkred", "FALSE" = "black"),
    guide = "none"  # hide color legend
  ) +
  
  theme_minimal() +
  labs(title = "Densiometer % Open by Direction", x = "Direction", y = "Percent Cover")
densiometer_box

# VIOLIN 
# 
# health_long <- health_long %>%
#   filter(direction != 'densio_average')

densiometer_violin <- health_long %>%
  ggplot(aes(x = direction, y = percent_cover)) +
  
  # Conditional fill and color
  geom_violin(
    aes(fill = direction == "densio_south", color = direction == "densio_south")
  ) +
  
  geom_jitter(width = 0.1, alpha = 0.5) +

  # Manual scales for TRUE/FALSE
  scale_fill_manual(
    values = c("TRUE" = "tomato", "FALSE" = "skyblue"),
    guide = "none"  # hide fill legend
  ) +
  scale_color_manual(
    values = c("TRUE" = "darkred", "FALSE" = "black"),
    guide = "none"  # hide color legend
  ) +

  theme_minimal() +
  labs(
    title = "Densiometer % Open by Direction",
    x = "Direction",
    y = "Percent Cover"
  )
densiometer_violin
```
### Individual Spagetti Plot
```{r}
seedlings_2025 <- health_assess_2025 %>% filter(is.na(dbh_cm))

seedling_health_long <- seedlings_2025 %>%
  mutate(case_id = row_number()) %>%
  pivot_longer(
    cols = starts_with("densio_"),
    names_to = "direction",
    values_to = "percent_cover"
  ) %>%
  mutate(
    quartile = ntile(percent_cover, 4)  # 4 = quartiles
  )


seedling_health_long <- seedling_health_long %>%
  mutate(line_type = "Individual")  # add a dummy column to map linetype

seedling_health_long <- seedling_health_long %>%
  filter(direction != "densio_average")

individual_spagetti_quartile <- seedling_health_long %>% ggplot(aes(x = direction, y = percent_cover, group = case_id)) +
  # Spaghetti lines with linetype
  geom_line(aes(linetype = line_type, color = factor(quartile)), alpha = 0.3) +

  # Points
  geom_point(aes(color = factor(quartile)), size = 2, alpha = 0.8) +

  # Mean line
  stat_summary(
    fun = mean,
    geom = "line",
    aes(group = 1, color = "Mean", linetype = "Mean"),
    size = 1.2
  ) +

  # Mean points
  stat_summary(
    fun = mean,
    geom = "point",
    aes(group = 1, color = "Mean"),
    size = 3
  ) +

  # Manual color and linetype scales
  scale_color_manual(
    name = "Legend",
    values = c(
      "1" = "#E41A1C",
      "2" = "#377EB8",
      "3" = "#4DAF4A",
      "4" = "#984EA3"
    ),
    labels = c("Q1", "Q2", "Q3", "Q4")
  ) +
  scale_linetype_manual(
    name = "Line Type",
    values = c("Individual" = "solid", "Mean" = "solid"),
    labels = c("Individual", "Mean")
  ) +

  theme_minimal() +
  labs(
    title = "% Open on Densiometer by Cardinal Direction",
    subtitle = "Thin lines represent individual seedlings across directions",
    x = "Cardinal Direction",
    y = "% Open on Densiometer"
  )
individual_spagetti
```


### North
```{r}
mean_north <- mean(health_assess_2025$densio_north, na.rm = TRUE)
median_north <- median(health_assess_2025$densio_north, na.rm = TRUE)

hist_densio_north <- health_assess_2025 %>% ggplot(aes(x = densio_north)) +
  geom_histogram() + # aes(fill = seedling_y_n)) +
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_north),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 32,
           y = 10, 
           label = paste("Mean =", round(mean_north, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_north),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 30,
           y = 13,
           label = paste("Median =", round(median_north, 2), "%"))

hist_densio_north

```

### South

```{r}
mean_south <- mean(health_assess_2025$densio_south, na.rm = TRUE)
median_south <- median(health_assess_2025$densio_south, na.rm = TRUE)

hist_densio_south <- health_assess_2025 %>% ggplot(aes(x = densio_south)) +
  geom_histogram() + # aes(fill = seedling_y_n)) +
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_south),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 32,
           y = 10, 
           label = paste("Mean =", round(mean_south, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_south),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 30,
           y = 13,
           label = paste("Median =", round(median_south, 2), "%"))

hist_densio_south

```
### East
```{r}
mean_east <- mean(health_assess_2025$densio_east, na.rm = TRUE)
median_east <- median(health_assess_2025$densio_east, na.rm = TRUE)

hist_densio_east <- health_assess_2025 %>% ggplot(aes(x = densio_east)) +
  geom_histogram() + # aes(fill = seedling_y_n)) +
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_east),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 32,
           y = 10, 
           label = paste("Mean =", round(mean_east, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_east),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 30,
           y = 13,
           label = paste("Median =", round(median_east, 2), "%"))

hist_densio_east

```

### West
```{r}
mean_west <- mean(health_assess_2025$densio_east, na.rm = TRUE)
median_west <- median(health_assess_2025$densio_east, na.rm = TRUE)

hist_densio_west <- health_assess_2025 %>% ggplot(aes(x = densio_west)) +
  geom_histogram() + # aes(fill = seedling_y_n)) +
  ylab("Number of Individuals") + 
  # Mean
  geom_vline(
    aes(xintercept = mean_west),
    color = "#bdbdbd",
    linetype = "dashed",
    size = 1
  ) +
  annotate("text",
           color = "#bdbdbd",
           x = 32,
           y = 10, 
           label = paste("Mean =", round(mean_west, 2), "%")) +
  
  # Median
  geom_vline(
    aes(xintercept = median_west),
    color = "#636363",
    linetype = "solid",
    size = 1
  ) +
  annotate("text",
           color = "#636363",
           x = 30,
           y = 13,
           label = paste("Median =", round(median_west, 2), "%"))

hist_densio_west

```
```{r}
library(patchwork)

empty <- plot_spacer()

(
  (empty +            hist_densio_west  + empty) /
  (hist_densio_west + empty +             hist_densio_east) /
  (empty +            hist_densio_south + empty))

  
```


# Summary Plots
```{r echo=FALSE, message=FALSE, warning=FALSE}
(bar_seedlings)

(hist_plant_height / hist_dbh)

bar_canker_severity

hist_percent_live_canopy_2025
  
#(hist_epicormic_base_2025 / hist_epicormic_trunk_2025 ) + plot_layout(guides="collect")

((bar_has_canker_2025 / bar_has_callous_2025))

((hist_trunk_canker / hist_base_canker))+plot_layout(guides="collect")

```

