---
title: "Data_Cleaning_May_2025"
author: "Hank Helmers"
date: "2025-08-27"
output: html_document
---

# Data Cleaning of May 2025 
This script organized the inconsistencies across the Butternut Health Assessment data collected using the [May 2025 Version](https://docs.google.com/document/d/15MV2UUqgxR9c2eEpp-V7zLcjA9GGJhMaZdKkgN2rZSM/edit?tab=t.0#heading=h.kjmvt9c5fvdf) of the form. 

The responses aligned with this version of the form are from Rows 2-17 and Dates June 5th and 6th, 2025. These are the focus of the cleaning.

Below you will find the cleaning steps necessary for that version of the form


```{r}
# Remove the columns we don't use
health_assess_2025 <- health_assess_2025 %>% select(
  ## Basic information
  -`Slope (degree)`
  -`Aspect (N, NE, E, etc)`
  
  ## Collections
  -`What did you collect?`,
  -`If VOUCHERS were collected, how many?`,
  -`If LEAVES ON SILICA FOR DNA were collected, how many?`,
  -`If CUTTINGS were collected, how many?`,
  -`If SEEDS were collected, how many?`,
  -`If other collections were made, please describe them here including the number collected.`
)
  
```


## Photos
* Seperate the first and last photo numbers
* Identify camera used
* Link to the related photo album (create an artificial column)

```{r}
# Focus in on the photo columns only
may_photos <- may_2025 %>% select( 
  Timestamp, site_name, plant_number,  # Identifiable info
  Camera, first_photo, last_photo      # Photo info
)

# Seperate first and last photo


```


