---
title: "2025_reu_hypothesis_2_dbh"
output: 
  html_document:
          toc: true
          toc_depth: 4
date: "2025-07-21"
---

# Hypothesis 2: Higher DBH will be correlated with a higher percentage of canker on the trunk and base

-   Variables: DBH→ canker_area_base/trunk
-   Mechanism: a large, healthy tree might fight off the canker for longer amount of time while small trees are the opposite

# Data Import

## Importing 2024 & 2025 Data

See the full code for the break down for the importing and cleaning open the documents of the same name. Or here: - [2024 cleaning code](https://rpubs.com/hankhelmers/cleaning_2024_health_form_data) and - [2025 cleaning code](https://rpubs.com/hankhelmers/cleaning_2025_health_form_data)

### Create the R files if needed.

Purl (or stitch) together R code from the markdown cleaning files. These stitched files are stored in 'purl' folder with the date of the code purled.

```{r purl_cleaning, EVAL=FALSE, IGNORE=TRUE}
# library(knitr)
# 
# purl("cleaning_2024_health_form_data.Rmd",
#      output =
#        paste("purl/cleaning_2024_health_form_data",
#        Sys.Date(),".R"))
# 
# purl("cleaning_2025_health_form_data.Rmd",
#      output =
#        paste("purl/cleaning_2025_health_form_data",
#        Sys.Date(),".R"))
```

### Run the R files to get cleaned data

Next, we can run those extracted R files to actually import and clean the data.

```{r message=FALSE, warning=FALSE}
## 2024 data
source(paste("purl/cleaning_2024_health_form_data", Sys.Date(), ".R"))

## 2025 data
source(paste("purl/cleaning_2025_health_form_data", Sys.Date(), ".R"))
```

### Combining 2024 and 2025 data along canker area & DBH

```{r combined_years, message=FALSE, warning=FALSE}
library(dplyr)

# Add a 'year' column to each dataset to allow grouping/plotting by year
health_assess_2024 <- health_assess_2024 %>%
  mutate(year = 2024)

health_assess_2025 <- health_assess_2025 %>%
  mutate(year = 2025)

# Keep only the necessary columns for this hypothesis
sliced_2024 <- health_assess_2024 %>%
  select(year, dbh_cm, base_canker_area, trunk_canker_area) 

sliced_2025 <- health_assess_2025 %>%
  select(year, dbh_cm, base_canker_area, trunk_canker_area) 

# Combine the two data frames
combined_2024_2025 <- bind_rows(sliced_2024, sliced_2025) 
```

### Select for Adults only due to DBH
```{r}
health_assess_2025 <- health_assess_2025 %>% filter(!is.na(dbh_cm))
```


# Setup of Analysis

This chunk defines a re-usable function to run a regression across multiple input datas.

```{r do_hypoth2_analysis, message=FALSE, warning=FALSE, COLLAPSE=TRUE}
library(tidyverse)
library(dplyr)

do_hypoth2_analysis <- function(data, predictor, response, xlabel, ylabel, color = "black") {
  # Define linear model
  model <- lm(reformulate(predictor, response), data = data)
  coefs <- coef(model)
  r_squared <- summary(model)$r.squared
  p_value <- summary(model)$coefficients[2,4]   # pull p.value from linear regression model
  # print(p_value)
  
  # Create equation and R^2 text
  equation <- paste0("y = ", round(coefs[2], 2), "x + ", round(coefs[1], 2))
  r2_text <- paste0("italic(R)^2 == ", round(r_squared, 3))
  p_text <- paste0("p-value = ", round(p_value, 3))
  
  # Get plotting ranges
  max_x <- max(data[[predictor]], na.rm = TRUE)
  
  max_y <- max(data[[response]], na.rm = TRUE)
  min_y <- min(data[[response]], na.rm = TRUE)
  range_y <- max_y - min_y
  
  # Annotate sample size near the bottom-right
  sample_size_y_position <- min_y + 0.08 * range_y
  
  # Annotate equation and r2 near the top-right
  p_y_position  <- max_y - 0.05 * range_y
  #eq_y_position <- max_y - 0.12 * range_y
  r2_y_position <- max_y - 0.15 * range_y
  
  # PLOT ------------------------------------------------------
  ggplot(data, aes_string(x = predictor, y = response)) +
    # Points
    geom_point(color = color, size = 3, alpha=0.75) +
    theme_bw() + 
    xlim(0, 80) + 
    ylim(0, 100) + 
    
    # Line of best fit
    #geom_smooth(method = "lm", color = color, se = F) +
    
    # Axes Label & Theme
    labs(x = xlabel, y = ylabel) +  # Set the y-axis label
    theme(
      axis.title.x = element_text(size = 16),
      axis.title.y = element_text(size = 16, color = color),
      axis.text = element_text(size = 14)
    ) +

    # Samples size text
    annotate(
      "text",
      x = max_x,
      y = sample_size_y_position,
      label = paste("n = ", count(data)),
      hjust = 1,
      size = 6,
      color = "black"
    ) + 

    # #Equation 
    # annotate(
    #   "text",
    #   x = max_x,
    #   y = eq_y_position,
    #   label = equation,
    #   hjust = 1,
    #   size = 4,
    #   color = color
    # ) +
    # R^2 text
    # annotate(
    #   "text",
    #   x = max_x,
    #   y = r2_y_position,
    #   label = r2_text,
    #   hjust = 1,
    #   size = 6,
    #   color = color,
    #   parse = TRUE # Ensures that italics styling works
    # ) +

    # P value text
    annotate(
      "text",
      x = max_x,
      y = p_y_position,
      label = p_text,
      hjust = 1,
      size = 6,
      color = color
    )
}
```

# Data Analysis

Here are the major delineations of questions across this hypotheses: \* Done ? In-progress

-   What patterns of DBH and canker areas do we see overall-- across all sites, years and age classes?
-   What patterns of DBH and canker areas do we see across years? \*
-   What patterns of DBH and canker areas do we see across sites? ?

Notably: \* I'd like to include both base and trunk canker area on the same graph; I think that could be very nice

## Patterns overall-- across all individuals (all sites, years, and age classes)

```{r message=FALSE, warning=FALSE}
library(scales)

hex <- hue_pal()(3)

patterns_across_all_base <- do_hypoth2_analysis(
  data = combined_2024_2025, 
  predictor = "dbh_cm", 
  response = "base_canker_area",
  xlabel = "All Individuals DBH (cm)",
  ylabel = "% of Base Canker Area",
  color = hex[1]
)
patterns_across_all_base


patterns_across_all_trunk <- do_hypoth2_analysis(
  data = combined_2024_2025, 
  predictor = "dbh_cm", 
  response = "trunk_canker_area",
  xlabel = "All Individuals DBH (cm)",
  ylabel = "% of Trunk Canker Area",
  color = hex[2]
)
patterns_across_all_trunk
```

## Patterns across years

### Patterns across 2025

```{r patterns_2025, message=FALSE, warning=FALSE}
library(scales)

hex <- hue_pal()(3)

adults_2025 <- health_assess_2025 %>% filter(!is.na(dbh_cm))

patterns_across_2025_base <- do_hypoth2_analysis(
  data = adults_2025, 
  predictor = "dbh_cm", 
  response = "base_canker_area",
  xlabel = "DBH (cm) in 2025 Adults",
  ylabel = "% of Base Canker Area",
  color = hex[1]
)
patterns_across_2025_base

patterns_across_2025_trunk <- do_hypoth2_analysis(
  data = adults_2025, 
  predictor = "dbh_cm", 
  response = "trunk_canker_area",
  xlabel = "Diameter at breast height (cm)",
  ylabel = "% of Trunk Canker Area",
  color = "#FF8E00"
)
patterns_across_2025_trunk

patterns_across_2025_girdled <- do_hypoth2_analysis(
  data = adults_2025, 
  predictor = "dbh_cm", 
  response = "girdled_canker_circum",
  xlabel = "2025 Individuals DBH (cm)",
  ylabel = "% of Circumference Girdled",
  color = hex[3]
)
patterns_across_2025_girdled
```

### Patterns across 2024

```{r patterns_2024, message=FALSE, warning=FALSE}
library(scales)

hex <- hue_pal()(3)

patterns_across_2024_base <- do_hypoth2_analysis(
  data = health_assess_2024, 
  predictor = "dbh_cm", 
  response = "base_canker_area",
  xlabel = "2024 Individuals DBH (cm)",
  ylabel = "% of Base Canker Area",
  color = hex[1]
)
patterns_across_2024_base

patterns_across_2024_trunk <- do_hypoth2_analysis(
  data = health_assess_2024, 
  predictor = "dbh_cm", 
  response = "trunk_canker_area",
  xlabel = "2024 Individuals DBH (cm)",
  ylabel = "% of Trunk Canker Area",
  color = hex[2]
)
patterns_across_2024_trunk
```


------------------------------------------------------------------------

# Results

## In 2024 alone, there were no correlations in either base or trunk canker areas and DBH.

```{r results_1, message=FALSE, warning=FALSE}
library(patchwork)

patterns_across_2024_base + patterns_across_2024_trunk
```


## In 2025 alone, there were no correlations in either base or trunk canker areas and DBH.

```{r results_2, message=FALSE, warning=FALSE}
patterns_across_2025_base + patterns_across_2025_trunk
```

## Across years: Appears to be an increase in DBH \~ Canker Area over 2024 and 2025.However, this could be a result of increasing our sample size almost two fold.

```{r results_4, message=FALSE, warning=FALSE}
(patterns_across_2024_base + patterns_across_2025_base) + plot_annotation("% Base Canker Area 2024 versus 2025")

(patterns_across_2024_trunk + patterns_across_2025_trunk) + plot_annotation("% Trunk Canker Area 2024 versus 2025")
```

## 2025 Girdled circumference: No correlation between canker area and DBH

```{r results_3, message=FALSE, warning=FALSE}
patterns_across_2025_girdled
```




# Future work

## Comparing base and trunk areas on same graph

## Two response variables (base/trunk compared together)

```{r eval=FALSE}
# 
# # This ensures adds a DBH measure to each of the canker entries, such that they can be plotted against each other. 
# data <- health_assess_2025
# predictor <- "dbh_cm"
# response1 <- "base_canker_area"
# 
# long_data <- data %>%
#   select(all_of(c(predictor, response1, response2))) %>%
#   pivot_longer(
#     cols = c(all_of(response1), all_of(response2)),
#     names_to = "response_type",
#     values_to = "response_value"
#   )
# 
# # Fit models and extract stats for annotation
# annotations <- long_data %>%
#   group_by(response_type) %>% # Collects all of the base/trunk canker area readings together
#   group_modify(~ {
#     model <- lm(response_value ~ get(predictor), data = long_health_data)
#     coefs <- coef(model)
#     r_squared <- summary(model)$r.squared
#     eq <- paste0("y == ", round(coefs[2], 2), " * x + ", round(coefs[1], 2))
#     r2 <- paste0("italic(R)^2 == ", round(r_squared, 4))
#     data.frame(equation = eq, r2 = r2)
#   })
# 
# # Determine top-right positions for annotation
#   max_x <- max(long_data[[predictor]], na.rm = TRUE)
#   max_y <- max(long_data$response_value, na.rm = TRUE)
#   min_y <- min(long_data$response_value, na.rm = TRUE)
#   range_y <- max_y - min(long_data$response_value, na.rm = TRUE)
# 
#   # Add y positions for annotations
#   # annotations["base_canker_area"]$eq_y <- 
#   annotations$eq_y <- min_y + 0.05 * range_y
#   annotations$r2_y <- min_y + 0.12 * range_y
#   annotations$x <- max_x
# 
#   # Plot
#   ggplot(long_data, aes_string(x = predictor, y = "response_value", color = "response_type")) +
#     geom_point() +
#     geom_smooth(method = "lm", se = FALSE) +
#     # xlab(xlabel) +
#     # ylab(ylabel) +
#     # Add annotations for each response type
#     geom_text(data = annotations,
#               aes(x = x, y = eq_y, label = equation, color = response_type),
#               hjust = 1, size = 4, parse = TRUE, inherit.aes = FALSE) +
#     geom_text(data = annotations,
#               aes(x = x, y = r2_y, label = r2, color = response_type),
#               hjust = 1, size = 4, parse = TRUE, inherit.aes = FALSE) +
#     theme_minimal()
# 
# view(annotations)
```
